---
title: "AKI Original (Paper-Style Reproduction)"
output: rmarkdown::html_vignette
css: styles.css
vignette: >
  %\VignetteIndexEntry{AKI Original (Paper-Style Reproduction)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-defaults, echo=FALSE, results=FALSE, message=FALSE}
source("defaults.R")
knitr::opts_chunk$set(fig.dim = c(7, 5), out.width = "50%", eval = TRUE)
```

## The Analysis Goal

This vignette follows the procedure described in Zacharias et al. (2012) to
classify AKI vs non-AKI patients based on 24h urine NMR spectra. The goal is to
reproduce the reported performance of 76% accuracy and 83% AUC.

## The Original Analysis Procedure

Relevant parts from the original paper are:

> The region from 9.5 to 0.5 ppm of the 1D spectra was evenly split into
> bins of 0.01. After exclusion of the solvent area and the area
> containing the broad urea signal from 6.5 to 4.5 ppm, 701 bins remained
> for each 1D spectrum.
>
> All spectra were normalized prior to analysis by Quantile Normalization
> (Bolstad et al. 2003; Kohl et al. 2012) to achieve equal distributions
> of feature intensities across spectra.
>
> Classification of spectra was performed using a SVM provided in the
> R-library e1071 (http://cran.r-project.org/ web/packages/e1071) within a
> nested cross-validation approach.
>
> For this contribution a SVM with a radial basis function kernel was used
> within a leave-five-out cross-validation.
>
> The internal parameters of the SVM were calibrated within an inner
> cross-validation loop employing a grid search procedure.

## Helper Functions

Based on these instructions we're gonna implement the following helper functions:

- `bin_spectra(spectra, ignore_region = c(6.5, 4.5))`

   For creating a matrix of binned features from a list of spectra, with
   optional ppm region exclusion.

- `get_quantile_reference(X)`

   For computing the reference distribution for quantile normalization from a
   feature matrix `X`.

- `quantile_normalize(X, ref)`

   For applying quantile normalization to a feature matrix `X` using a reference
   distribution `ref`.

- `train_svm(spectra, cls, cost = 0.5, gamma = 2^-10, nfeat=3, normalize = TRUE)`

  For training a radial SVM classifier with margin regularization strength
  `cost`, RBF kernel width `gamma` and `nfeat` number of features. Features are
  obtained by binning the given input spectra using `bin_spectra()` and
  selecting the top `nfeat` based on t-scores from a Welch t-test, comparing the
  two classes in `cls`. If `normalize` is TRUE, quantile normalization is
  applied to the binned features before feature selection and SVM training. The
  feature-selection based on t-scores is described in 'Gronwald et al., 2011'.
  The returned object will contain the trained SVM model, the
  quantile-normalisation reference distribution and the indices of the selected
  features.

- `find_best_params(X, y, costs, gammas, nfeats, normalize, k = 5)`

  For training multiple SVMs across a grid using k-fold cross-validation to
  select the best parameters. In this vignette we use `costs = (1:10) / 10`,
  `gammas = 2^(-12:-5)` and `nfeats = 2:3`. We limit the number of features to
  such a narrow range, because the original paper states, that the average
  optimum is 2.4 across validation folds with a standard deviation of 0.5. This
  means, that only values of 2 and 3 can have been observed (e.g. 2,2,2,3,3) or
  the SD would have been higher. Since the goal of this article is not to rerun
  the whole analysis workflow, but to reproduce the reported results as closely
  as possible, there is no need to search a wider range of feature counts.

## The Final Classification Pipeline

The final classification pipeline will then be implemented as follows:

```{r pipeline, echo=TRUE}
benchmark_aki_classification <- function() {

   xds_dir <- download_example_datasets()
   path <- file.path(xds_dir, "bruker", "aki")

    spectra <- read_spectra(path)
    meta <- read_metadata(path)

    X <- bin_spectra(spectra)
    y <- meta$HasAKI
    n <- nrow(meta)
    k <- 5

    train_ids <- get_train_ids(k, n)
    test_ids <- get_test_ids(k, n)

    results <- lapply(seq_len(k), function(i) {
        X_train <- X[train_ids[[i]], ]
        y_train <- y[train_ids[[i]]]
        X_test <- X[test_ids[[i]], ]
        y_test <- y[test_ids[[i]]]

        param_grid <- find_best_params(X_train, y_train)
        idx_best <- which.max(param_grid$auc)
        cost <- param_grid[idx_best, "cost"]
        gamma <- param_grid[idx_best, "gamma"]
        nfeat <- param_grid[idx_best, "nfeat"]

        svm <- train_svm(X_train, y_train, cost, gamma, nfeat)

        y_pred <- predict(svm, X_test)
        auc <- auc(y_test, y_pred)
        acc <- mean(y_test == y_pred)

        list(cost, gamma, nfeat, auc, acc)
    })

    results <- do.call(rbind, results)
    colnames(results) <- c("cost", "gamma", "nfeat", "auc", "acc")
    rownames(results) <- paste0("Fold", seq_len(k))
    means <- colMeans(results)
    sds <- apply(results, 2, sd)
    rbind(results, Mean = means, SD = sds)

    results # 7 * 5 data.frame with fold-wise and summary performance
}
```

## Step by Step Execution

To make the analysis as transparent as possible, we will now execute the above
pipeline step by step, visualizing intermediate results and outputs along the
way, before showing the final results at the end of the vignette.

### Downloading the Data

The first thing we need to do is to download the data, if not already available
locally. This is done via `download_example_datasets()`. The AKI dataset is
then available in the `bruker/aki` subdirectory of the downloaded data. After
that we will use metabodecon's `tree()` function to visualize the directory
structure of the downloaded data.

```
xds_dir <- metabodecon::download_example_datasets()
aki_dir <- file.path(xds_dir, "bruker", "aki")
metabodecon::tree(aki_dir)
```