---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
figure, div.figure
{
    border: 1px solid #dee2e6;
    border-radius: .375rem;
    padding: 0;
    margin: 1em 0;
}
caption, p.caption {
    text-align: center;
    font-style: italic;
    border-top: 1px solid #dee2e6;
    padding: 0.75rem;
    background-color: #ffffff;
}
img {
    margin: 0 auto;
    padding: 0;
    border: none;
    width: 50%;
}
@media (max-width: 500px) {
    img {
        width: 100%; /* Full width for smaller viewports */
    }
}
</style>

```{r set-defaults, echo=FALSE, results=FALSE, message=FALSE}
knitr::opts_chunk$set(
    dev = "svg",
    fig.dim = c(5, 5), # Dimensions of stores figures in inches
    fig.show = "hold", # Render images as inline elements
    out.height = "auto", # (1)
    eval = FALSE,
    echo = FALSE,
    results = FALSE,
    message = FALSE
    # (1) Either out.width or out.heigt must be set of rmarkdown will not put
    # a div.figure around the individual img elements
)
```

This article shows how Metabodecon can be used for  deconvoluting  and  aligning
one-dimensional NMR spectra using the pre-installed [Sim] dataset as an example.
The Sim dataset includes 16  simulated  spectra,  each  with  2048  data  points
ranging from ≈ 3.6 to  3.3  ppm.  These  simulated  spectra  closely  mimic  the
resolution and signal strength of real NMR experiments on blood plasma  from  16
individuals. The Sim dataset is used instead of the Blood dataset because it  is
smaller, faster to process, and comes pre-installed with the package.  For  more
information on the Sim and Blood datasets, see [Datasets].

[Sim]: https://spang-lab.github.io/metabodecon/articles/Datasets.html#sim
[Datasets]: https://spang-lab.github.io/metabodecon/articles/Datasets.html

# Deconvolute spectra

To find the path to the  Sim  dataset,  you  can  use  the  `metabodecon_file()`
function, which returns the path to any file or  directory  within  the  package
directory. To deconvolute the  spectra  within  the  Sim  dataset  you  can  use
`generate_lorentz_curves()`:

```{r chunk-deconvolute, echo=TRUE}
sim_dir <- metabodecon::metabodecon_file("bruker/sim")
deconvoluted_spectra <- metabodecon::generate_lorentz_curves(
    data_path = sim_dir, # Path to directory containing spectra
    sfr = c(3.35, 3.55), # Borders of signal free region (SFR) in ppm
    wshw = 0,            # Half width of water signal (WS) in ppm
    smopts = c(2, 5),    # Configure smoothing parameters
    verbose = FALSE      # Suppress status messages
)
```

After calling `generate_lorentz_curves()`, the function will ask you  to  answer
the following questions to determine the optimal deconvolution parameters:

1. Use same parameters for all spectra? (y/n)
2. Number of spectrum for adjusting parameters? (1: sim_01, 2: sim_02, ...)
3. Signal free region correctly selected? (y/n)
4. Water artefact fully inside red vertical lines? (y/n)

You can answer questions one and two  with  `y`  and  `1`,  as  the  dataset  is
homogeneous, i.e., all spectra were measured in  the  same  lab  with  the  same
acquisition and processing parameters. However, for heterogeneous datasets, it's
advisable to optimize parameters for each batch of spectra individually.

Questions three and  four  are  accompanied  by  two  plots,  shown  in  [Figure
1](#fig-deconvolute), which help you to verify  the  accuracy  of  the  selected
signal-free region (SFR) and water signal half-width  (WSHW)  [^wshw].  In  this
case, the provided parameters are already fine, so you can answer both questions
with `y`. If adjustments are needed, you can respond  with  `n`  and  input  the
correct values.

<a id="fig-deconvolute"></a>

```{r fig-deconvolute, eval=TRUE, fig.cap='<strong>Figure 1.</strong> The first spectrum of the Sim dataset. The x-Axis gives the chemical shift of each datapoint in parts per million (ppm). The y-Axis gives the signal intensity of each datapoint in arbitrary units (au). The signal free regions are shown as green rectangles in the left plot. The water signal region, usually shown as blue rectangle, is shown in the right plot. Because the water signal half width is set to zero in this case, the rectangle collapses to a vertical line.'}
metabodecon::evalwith(
    answers = list(sameParams = "y", adjNo="1", sfrOk = "y", wsOk = "y"),
    pars = list(mar = c(4, 4, 2, 2)),
    expr = {
        <<chunk-deconvolute>>
    }
)
```

When using the function in scripts, where interactive user input is not desired,
you can disable the interactive prompting by setting parameter `ask` to `FALSE`.
In this case, the provided parameters will be used for the deconvolution of all
spectra automatically. [^askFALSE]

# Visualize deconvoluted spectra

After completing the deconvolution, it is advisable to visualize the extracted
signals using `plot_spectrum()` to assess the quality of the deconvolution.

```{r chunk-plot-spectrum, echo=TRUE}
# Visualize the first spectrum.
metabodecon::plot_spectrum(deconvoluted_spectra[[1]])

# Visualize the second spectrum, this time without the legend.
metabodecon::plot_spectrum(deconvoluted_spectra[[1]], lgd = FALSE)

# Visualize all spectra and save them to a pdf file
pdfpath <- tempfile(fileext = ".pdf")
pdf(pdfpath)
for (x in deconvoluted_spectra) {
    metabodecon::plot_spectrum(x, main = x$filename)
}
dev.off()
cat("Plots saved to", pdfpath, "\n")
```

Out of the 16 generated plots, the first two are shown as  examples  in  [Figure
2](#fig-plot-spectrum). Things to look out for are:

1. That the smoothing does not remove any real signals. If the smoothing is  too
   strong, i.e., the smoothed signal intensity (SI) is very different  from  the
   raw SI, you should adjust the smoothing parameters `smopts` in  the  call  to
   `generate_lorentz_curves()`.
2. That   the superposition of the lorentz curves is a good approximation of the
   smoothed SI. If major peaks are missed by the algorithm,  you  should  reduce
   the threshold `delta` in the call to `generate_lorentz_curves()`.

<a id="fig-plot-spectrum"></a>

```{r fig-plot-spectrum, eval=TRUE, fig.cap="<strong>Figure 2.</strong> Deconvolution results for the first two spectra of the Sim dataset. The raw SI (black), smoothed SI (blue), and superposition of Lorentz curves (red) are closely aligned, indicating that <code>smopts</code> and <code>delta</code> were chosen well and that the deconvolution was successful."}
<<chunk-plot-spectrum>>
```

# Align deconvoluted spectra

The last step in the Metabodecon Workflow is to align the deconvoluted  spectra.
This is necessary because the chemical shifts of the peaks in  the  spectra  may
vary slightly due to differences in the measurement conditions.

To perform the alignment, you can use `align()`. To visualize  the  data  before
and after the alignment, you can use `plot_spectra()`:

```{r chunk-align, echo=TRUE}
# Plot spectra before alignment. Only shows spectra 1-8 for clarity.
metabodecon::plot_spectra(deconvoluted_spectra[1:8], lgd = FALSE)

# Align spectra and plot again.
aligned_spectra <- metabodecon::align(deconvoluted_spectra)
metabodecon::plot_spectra(aligned_spectra[1:8])
```

The resulting plots are shown in [Figure 3](#fig-align). Before  the  alignment,
the spectra exhibit generally similar shapes but do not perfectly overlap. After
the alignment, the spectra are much more consistent with each other,  indicating
that the alignment was  successful.  Notably,  spectrum  two  has  been  shifted
significantly to the left.

<a id="fig-align"></a>
```{r fig-align, eval=TRUE, fig.cap='<strong>Figure 3.</strong> Overlay of the first eight deconvoluted spectra from the Sim dataset before alignment (left) and after alignment (right). The x-Axis gives the chemical shift of each datapoint in parts per million (ppm). The y-Axis gives the signal intensity of each datapoint in arbitrary units (au). All specta are pretty similar to each other except for Spectrum 2, which got shifted approx. 0.01 ppm to the right.'}
<<chunk-align>>
```

<!-- Footnotes -->

[^wshw]: Since the used dataset was simulated based on  actual  measurements  in
    the  range  of  approx.  3.6  to  3.3  ppm,  it   doesn't  contain  a  water
    signal. Therefore, the water signal half width was set to zero.

[^askFALSE]: In this case it may be useful to set `ask = TRUE` for the first run
    to determine the optimal parameters interactively and then set `ask = FALSE`
    for subsequent runs.
