---
title: "Metabodecon Models"
output:
  rmarkdown::html_vignette: default
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{fontspec}
  - \usepackage{etoolbox}
  - \usepackage{fvextra}
  - \usepackage{needspace}
  - \definecolor{shadecolor}{RGB}{232,232,232}
  - \fvset{breaklines=true,breakanywhere=true}
  - \BeforeBeginEnvironment{Shaded}{\Needspace{10\baselineskip}}
  - \BeforeBeginEnvironment{Shaded}{\vspace{0.5em}}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines=true,breakanywhere=true,commandchars=\\\{\}}
  - \RecustomVerbatimEnvironment{verbatim}{Verbatim}{breaklines=true,breakanywhere=true}
css: styles.css
vignette: >
  %\VignetteIndexEntry{Metabodecon Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-defaults, echo=FALSE, results=FALSE, message=FALSE}
knitr::opts_chunk$set(
    fig.dim=c(5, 5), fig.show="hold", out.width="50%",
    echo=TRUE, message=FALSE, warning=FALSE
)
```

This article demonstrates how to use `fit_mdm()` and `benchmark()` to build,
tune and evaluate classification models on deconvoluted NMR spectra. We use
the simulated `sim2` dataset shipped with the package, split it into training
and test sets, run a small grid search to find good preprocessing parameters,
and then evaluate on held-out data.

# Load spectra

`sim2` contains 100 simulated 1D NMR spectra split evenly into groups `A` and
`B`. Five out of 25 peaks per spectrum differ between groups by 10% in area.
The group labels are attached as an attribute. For details on how the dataset
was constructed, see `?sim2`.

```{r load-spectra}
library(metabodecon)
x <- sim2
y <- attr(sim2, "group")
n <- length(x)
```

The following figure shows four spectra from each group overlaid on top of each
other. The subtle area differences are hard to spot visually.

```{r plot-simulated}
a4 <- which(y == "A")[1:4]
b4 <- which(y == "B")[1:4]
ab <- c(a4, b4)
plot_spectra(x[ab])
heat_spectra(x[ab])
```

# Train/test split

We use one half of the data for training and one half for testing.

```{r split}
set.seed(1)
n_train <- round(0.5 * n)
tr <- sort(sample(n, n_train))
te <- setdiff(seq_len(n), tr)
```

# Fit a model manually, step by step

To go from raw spectra to a fitted classifier we need three preprocessing
stages plus the model fit itself:

1. **Deconvolution** -- represent each spectrum as a list of Lorentzian peaks.
2. **Alignment** -- shift peaks so corresponding peaks across spectra share the
   same chemical-shift index. One spectrum is picked as the *reference* and
   all others are aligned to it.
3. **Feature matrix construction** -- collapse the aligned peak lists into one
   row per spectrum, optionally merging peaks within `maxCombine` indices.
4. **Model fit** -- here, an L1-penalized logistic regression via
   `cv.glmnet()`.

`fit_mdm()` performs all four steps for every row of a parameter grid and
returns the best model. Below we do them by hand for one parameter set so we
can inspect each intermediate state.

We pick a small group-balanced subset (4+4 spectra) just for plotting; the
model itself is trained on all of `x[tr]`.

```{r fit-manually-prep}
a4tr <- which(y[tr] == "A")[1:4]
b4tr <- which(y[tr] == "B")[1:4]
abtr <- c(a4tr, b4tr)
yab <- y[tr][abtr]
true_x0 <- attr(sim2, "true_x0")  # post-alignment ppm of discriminating peaks
```

## Step 1: Deconvolute

```{r fit-manually-decon}
decons <- deconvolute(x[tr], nfit=10, smit=2, smws=5, delta=10, npmax=0, verbose=FALSE)
plot_spectra(decons[abtr])
heat_spectra(decons[abtr], y=yab)
```

## Step 2: Align

`clupa()` calls `find_ref()` internally to pick the spectrum whose
peaks are closest to all others. We extract that reference now -- we need it
later to align the *test* spectra to the same coordinate system.

```{r fit-manually-align}
aligns <- clupa(decons, maxShift=50, verbose=FALSE)
ref <- find_ref(decons)
plot_spectra(aligns[abtr])
heat_spectra(aligns[abtr], y=yab)
```

## Step 3: Build the feature matrix

`si_mat()` turns each aligned peak list into a row of a peak-area matrix.
Columns are aligned chemical-shift indices, optionally merged within
`maxCombine` indices of each other. `peakPos` records which columns are
actually populated -- these are the features the model will see.

```{r fit-manually-simat}
mat <- si_mat(aligns, maxCombine=10)
peakPos <- which(colSums(mat != 0) > 0)
X <- mat[, peakPos, drop=FALSE]
dim(X)
heat_spectra(X, y=y[tr], true_x0=true_x0)
heat_spectra(X, y=y[tr], true_x0=true_x0, scale_cols=TRUE)
```

The standardized view (`scale_cols=TRUE`) makes the group structure of the
discriminating peaks more visible: columns near `true_x0` show consistent
sign differences between A (top) and B (bottom).

## Step 4: Fit the lasso

```{r fit-manually-glmnet}
foldid <- metabodecon:::get_foldid(y=y[tr], nfolds=5, seed=1)
cvfit <- glmnet::cv.glmnet(X, y[tr], family="binomial", alpha=1, foldid=foldid, keep=TRUE)
plot(cvfit)
```

The non-zero coefficients at `lambda.min` are the chemical-shift columns the
model relies on. Let us highlight them in the feature heatmap:

```{r fit-manually-picked}
cf <- as.matrix(coef(cvfit, s="lambda.min"))
picked <- rownames(cf)[cf[, 1] != 0]
picked <- setdiff(picked, "(Intercept)")
picked_idx <- match(picked, colnames(X))
heat_spectra(X[, picked_idx, drop=FALSE], y=y[tr], true_x0=true_x0)
heat_spectra(X[, picked_idx, drop=FALSE], y=y[tr], true_x0=true_x0,
             scale_cols=TRUE)
```

## Equivalent one-shot call

The exact same model can be fit with one `fit_mdm()` call:

```{r fit-default}
gd <- data.frame(nfit=10, smit=2, smws=5, delta=10, npmax=0,
                 maxShift=50, maxCombine=5)
md <- fit_mdm(x[tr], y[tr], mog=gd, verbosity=0, nworkers=1, nfolds=5)
print(md)
```

# Predict held-out spectra manually

`predict.mdm()` mirrors the four training steps for new data, but reuses the
training-time reference and `peakPos` so the test feature matrix has the
**same columns** as the training one.

## Step 1: Deconvolute the test spectra

```{r predict-manually-decon}
decons_te <- deconvolute(x[te], nfit=10, smit=2, smws=5, delta=10,
                         npmax=0, verbose=FALSE)
```

## Step 2: Align to the training reference

```{r predict-manually-align}
aligns_te <- clupa(decons_te, maxShift=50, ref=ref, verbose=FALSE)
```

## Step 3: Build feature matrix using training `peakPos`

`si_mat(..., peakPos=peakPos)` ensures the test matrix has columns at the
training-time chemical-shift indices, then we restrict to the `peakPos`
subset to match `X`.

```{r predict-manually-simat}
mat_te <- si_mat(aligns_te, maxCombine=10, peakPos=peakPos)
Xte <- mat_te[, peakPos, drop=FALSE]
stopifnot(identical(colnames(Xte), colnames(X)))
heat_spectra(Xte, y=y[te], true_x0=true_x0, scale_cols=TRUE)
```

## Step 4: Apply the trained model

```{r predict-manually-glmnet}
prob_te <- as.numeric(predict(cvfit, newx=Xte, s="lambda.min", type="response"))
pred_te <- factor(ifelse(prob_te > 0.5, levels(y)[2], levels(y)[1]),
                  levels=levels(y))
acc_manual <- mean(pred_te == y[te])
auc_manual <- metabodecon:::AUC(y[te], prob_te)
cat(sprintf("Manual test acc: %.1f%%, AUC: %.3f\n", acc_manual * 100, auc_manual))
```

# Compare with a random forest

The feature matrix contains many zeros, which represent peaks that were not
detected in a given spectrum -- semantically closer to "missing" than to
"low intensity". Tree-based models handle this naturally: a single split at
`x>0` separates absent from present, and further splits on the present
subset capture intensity differences. Random forests also have built-in
regularization via tree depth and ensembling, so they need almost no tuning.

We fit a `ranger` random forest on the same `X`/`Xte` matrices used above:

```{r ranger-fit, eval=requireNamespace("ranger", quietly=TRUE)}
rf <- ranger::ranger(x=X, y=y[tr], probability=TRUE, num.trees=500, seed=1)
prob_rf <- predict(rf, data=Xte)$predictions[, levels(y)[2]]
pred_rf <- factor(ifelse(prob_rf > 0.5, levels(y)[2], levels(y)[1]), levels=levels(y))
acc_rf <- mean(pred_rf == y[te])
auc_rf <- metabodecon:::AUC(y[te], prob_rf)
cat(sprintf("RF test acc: %.1f%%, AUC: %.3f\n", acc_rf * 100, auc_rf))
```

The equivalent one-shot call using `fit_mdm()` with `model="ranger500"` runs
the same OOB-evaluated random forest over any preprocessing grid:

```{r ranger-fitmdm, eval=requireNamespace("ranger", quietly=TRUE)}
gd <- data.frame(nfit=10, smit=2, smws=5, delta=10, npmax=0,
                 maxShift=50, maxCombine=5)
md_rf <- fit_mdm(x[tr], y[tr], mog=gd, model="ranger500",
                 verbosity=0, nworkers=1, nfolds=5)
preds_rf <- predict(md_rf, x[te], type="all", verbosity=0)
acc_rf2 <- mean(preds_rf$class == y[te])
auc_rf2 <- metabodecon:::AUC(y[te], preds_rf$prob)
cat(sprintf("RF (fit_mdm) test acc: %.1f%%, AUC: %.3f\n", acc_rf2*100, auc_rf2))
```

# Tune preprocessing via grid search

A multi-row `mog` makes `fit_mdm()` evaluate a grid of preprocessing parameter
combinations. For each row it builds the feature matrix, runs `cv.glmnet()`
with a fixed fold assignment, and records the held-out accuracy and AUC at
`lambda.min`. We use a deliberately small grid here to keep vignette build
time short:

```{r tune-mdm}
gs <- expand.grid(smit=1:2, smws=c(3,5), delta=c(3.2,6.4), nfit=5, npmax=0, maxShift=50, maxCombine=5)
mt <- fit_mdm(x=x[tr], y=y[tr], mog=gs, nfolds=5, verbosity=1, nworkers=5)
```

The performance grid shows how each parameter combination scored:

```{r mog-results}
G <- mt$mog
G <- G[order(-G$auc),]
cap <- "AUC per Parameter Combination. Higher is better."
knitr::kable(x=head(G,10), row.names=FALSE, caption=cap)
```

# Compare deconvolution results

Let us compare the deconvolution of the first training spectrum using the worst
and the best parameters found by the grid search:

```{r compare-decon}
# Deconvolute first two training spectra with default and tuned parameters
w <- G[nrow(G),]
b <- G[1, ]
dw <- deconvolute(x[tr[1:2]], nfit=w$nfit, smit=w$smit, smws=w$smws, delta=w$delta, npmax=w$npmax, verbose=FALSE)
db <- deconvolute(x[tr[1:2]], nfit=b$nfit, smit=b$smit, smws=b$smws, delta=b$delta, npmax=b$npmax, verbose=FALSE)
plot_spectrum(db[[1]], main="Best params", foc_frac=0:1, lgd=FALSE)
plot_spectrum(dw[[1]], main="Worst params", foc_frac=0:1)
```

# Inspect lasso features

The non-zero coefficients of the lasso model reveal which chemical shift
positions are most important for distinguishing the two groups:

```{r coefs}
cf <- coef(mt)
nz <- cf[cf[, 1] != 0, , drop=FALSE]
knitr::kable(
    data.frame(feature=rownames(nz), coefficient=round(nz[, 1], 4)),
    row.names=FALSE,
    caption="Non-zero lasso coefficients at lambda.min."
)
```

# Predict test samples

We use `predict.mdm()` to classify the held-out test spectra. This internally
deconvolutes and aligns them against the reference spectrum stored in the model:

```{r predict-test}
preds <- predict(mt, x[te], type="all", verbosity=0)
results <- data.frame(
    sample=get_names(x[te]),
    true=y[te],
    predicted=preds$class,
    prob_B=round(preds$prob, 3)
)
```

```{r confusion-matrix}
cap <- "Confusion matrix on test set."
acc <- mean(preds$class == y[te])
auc <- metabodecon:::AUC(y[te], preds$prob)
knitr::kable(table(True=y[te], Pred=preds$class), caption=cap)
cat(sprintf("\nTest Acc: %.1f%%\n", acc * 100))
cat(sprintf("Test AUC: %.1f%%\n", auc * 100))
cat(sprintf("nA/nA+nB: %.1f%%\n", sum(y[te] == "A") / length(te) * 100))
```

# Benchmark with nested cross-validation

If you do not have enough samples for a train/test split, you can use
`benchmark()` instead. It wraps `fit_mdm()` in an outer cross-validation loop
so that every sample is used for both training and evaluation. The outer
held-out fold is never seen during preprocessing or model fitting, so the
resulting accuracy and AUC are honest estimates.

A typical call looks like this:

```{r benchmark-code, eval=FALSE}
bm <- benchmark(x, y, fun="fit_mdm", mog=mog)
acc <- mean(bm$predictions$true == bm$predictions$pred) * 100
cat(sprintf("Nested CV accuracy: %.1f%%\n", acc))
```

Note that `benchmark()` is computationally expensive because it runs the full
grid search for each outer fold. It is therefore not executed in this
vignette.
