---
title: "Metabodecon Models on SAP2"
output:
  rmarkdown::html_vignette: default
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{fontspec}
  - \usepackage{etoolbox}
  - \usepackage{fvextra}
  - \usepackage{needspace}
  - \definecolor{shadecolor}{RGB}{232,232,232}
  - \fvset{breaklines=true,breakanywhere=true}
  - \BeforeBeginEnvironment{Shaded}{\Needspace{10\baselineskip}}
  - \BeforeBeginEnvironment{Shaded}{\vspace{0.5em}}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines=true,breakanywhere=true,commandchars=\\\{\}}
  - \RecustomVerbatimEnvironment{verbatim}{Verbatim}{breaklines=true,breakanywhere=true}
css: styles.css
vignette: >
  %\VignetteIndexEntry{Metabodecon Models on SAP2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-defaults, echo=FALSE, results=FALSE, message=FALSE}
knitr::opts_chunk$set(
    fig.dim=c(5, 5), fig.show="hold", out.width="50%",
    echo=TRUE, message=FALSE, warning=FALSE
)
```

This article walks through the full metabodecon pipeline -- deconvolution,
alignment, feature extraction, model fitting and prediction -- on the
[`sap2`][metabodecon::sap2] mini classification dataset. `sap2` is small
enough (18 spectra x 128 datapoints) that we can plot every spectrum, every
peak list and every feature column in full, without hiding anything behind
zoomed views or column subsets. This makes it useful for visualizing what
each pipeline stage actually does. For a more realistic but bigger example,
see the [Metabodecon Models](MDM.html) vignette which uses the
[`sim2`][metabodecon::sim2] dataset.

# Load spectra

`sap2` contains 18 simulated 1D NMR spectra split evenly into groups `A` and
`B`. Each spectrum has 4 peaks; one of them differs between groups by 10%
in area. Group labels are attached as an attribute. For details on how the
dataset was constructed, see `?sap2`.

```{r load-spectra}
library(metabodecon)
x <- sap2
y <- attr(sap2, "group")
n <- length(x)
true_x0 <- attr(sap2, "true_x0") # ppm of the discriminating peak
```

The full set of raw spectra fits in two plots: a line overlay coloured by
group, and a heatmap with one row per spectrum and one column per
datapoint:

```{r plot-raw}
plot_spectra(x)
heat_spectra(x, y=y)
```

# Train/test split

We split 9/9 train/test, keeping group balance:

```{r split}
set.seed(1)
tr <- sort(c(sample(which(y == "A"), 4), sample(which(y == "B"), 5)))
te <- setdiff(seq_len(n), tr)
```

# Manual pipeline, step by step

The [Metabodecon Models](MDM.html) vignette describes each stage in detail.
Here we only highlight the parts that look different on `sap2`.

The default deconvolution settings are tuned for ~2000-datapoint spectra and
do not work on `sap2`'s 128-datapoint window. We pass tighter parameters
explicitly: `sfr=c(3.58, 3.45)` defines the signal-containing region (so
~9 datapoints on each edge are treated as signal-free for noise estimation)
and `smit=0, smws=3, delta=4` use minimal smoothing and a higher detection
threshold to suppress spurious noise peaks.

## Step 1: Deconvolute

```{r decon}
dpar <- list(sfr=c(3.58, 3.45), smit=0, smws=3, delta=4, nfit=3, npmax=0)
decons <- do.call(deconvolute, c(list(x=x[tr], verbose=FALSE), dpar))
plot_spectra(decons)
heat_spectra(decons, y=y[tr])
```

## Step 2: Align

`clupa()` shifts peaks so corresponding peaks across spectra share the
same chemical-shift index. The reference spectrum picked here is reused
later when we predict on the test set. Because the per-spectrum global
shift in `sap2` is small (sd ~4 datapoints), `maxShift=4` is enough.

```{r align}
aligns <- clupa(decons, maxShift=4, verbose=FALSE)
ref <- find_ref(decons)
plot_spectra(aligns)
heat_spectra(aligns, y=y[tr])
```

## Step 3: Build the feature matrix

`si_mat()` collapses the aligned peak lists into one row per spectrum. The
full output has one column per chemical-shift index (128 here), most of
which are empty. `peakPos` records which columns are actually populated --
these are the features the model will see. Because both the full matrix
and the populated subset are small, we can visualize them in full:

```{r simat}
mat <- si_mat(aligns, maxCombine=2)
peakPos <- which(colSums(mat != 0) > 0)
X <- mat[, peakPos, drop=FALSE]
dim(X)
heat_spectra(mat, y=y[tr], true_x0=true_x0)
heat_spectra(X, y=y[tr], true_x0=true_x0)
heat_spectra(X, y=y[tr], true_x0=true_x0, scale_cols=TRUE)
```

## Step 4: Fit the lasso

```{r glmnet}
foldid <- metabodecon:::get_foldid(y=y[tr], nfolds=3, seed=1)
cvfit <- glmnet::cv.glmnet(X, y[tr], family="binomial", alpha=1,
                           foldid=foldid, keep=TRUE)
plot(cvfit)
```

The non-zero coefficients at `lambda.min` highlight the chemical-shift
columns the model relies on:

```{r picked}
cf <- as.matrix(coef(cvfit, s="lambda.min"))
picked <- setdiff(rownames(cf)[cf[, 1] != 0], "(Intercept)")
if (length(picked) > 0) {
    picked_idx <- match(picked, colnames(X))
    heat_spectra(X[, picked_idx, drop=FALSE], y=y[tr], true_x0=true_x0,
                 scale_cols=TRUE)
} else {
    cat("Lasso picked no non-zero features at lambda.min.\n")
}
```

# Predict held-out spectra

`predict.mdm()` mirrors steps 1-3 on new data using the training-time
reference and `peakPos`. Done manually:

```{r predict-manual}
decons_te <- do.call(deconvolute, c(list(x=x[te], verbose=FALSE), dpar))
aligns_te <- clupa(decons_te, maxShift=4, ref=ref, verbose=FALSE)
mat_te <- si_mat(aligns_te, maxCombine=2, peakPos=peakPos)
Xte <- mat_te[, peakPos, drop=FALSE]
heat_spectra(Xte, y=y[te], true_x0=true_x0, scale_cols=TRUE)

prob_te <- as.numeric(predict(cvfit, newx=Xte, s="lambda.min", type="response"))
pred_te <- factor(ifelse(prob_te > 0.5, levels(y)[2], levels(y)[1]),
                  levels=levels(y))
acc <- mean(pred_te == y[te])
auc <- metabodecon:::AUC(y[te], prob_te)
cat(sprintf("Manual test acc: %.1f%%, AUC: %.3f\n", acc * 100, auc))
```

# Equivalent one-shot call with `fit_mdm()`

The same model can be fit with one `fit_mdm()` call. We pass the custom
`sfr` directly and set the per-row preprocessing parameters via `mog`:

```{r fitmdm}
gd <- data.frame(nfit=3, smit=0, smws=3, delta=4, npmax=0, maxShift=4, maxCombine=2)
md <- fit_mdm(x[tr], y[tr], mog=gd, sfr=c(3.58, 3.45), verbosity=0, nworkers=1, nfolds=3)
print(md)
```

```{r fitmdm-predict}
preds <- predict(md, x[te], type="all", verbosity=0)
acc <- mean(preds$class == y[te])
auc <- metabodecon:::AUC(y[te], preds$prob)
cat(sprintf("fit_mdm test acc: %.1f%%, AUC: %.3f\n", acc * 100, auc))
```

# Caveats

`sap2` is intentionally tiny. With only 9 training samples, 4 peaks per
spectrum and a 10% effect size on a single discriminator, lasso has very
little signal to latch on to: held-out accuracy is close to chance and
varies substantially with the train/test split. The point of this dataset
is to make every pipeline stage *visible* end-to-end, not to demonstrate
strong predictive performance. For the latter, use [`sim2`][metabodecon::sim2]
or a real dataset.
