---
title: "AKI Original (Paper-Style Reproduction)"
output:
  rmarkdown::html_document: default
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{fontspec}
  - \usepackage{etoolbox}
  - \usepackage{fvextra}
  - \usepackage{needspace}
  # - \setmonofont{Menlo}
  - \definecolor{shadecolor}{RGB}{232,232,232}
  - \fvset{breaklines=true,breakanywhere=true}
  - \BeforeBeginEnvironment{Shaded}{\Needspace{10\baselineskip}}
  - \BeforeBeginEnvironment{Shaded}{\vspace{0.5em}}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines=true,breakanywhere=true,commandchars=\\\{\}}
  - \RecustomVerbatimEnvironment{verbatim}{Verbatim}{breaklines=true,breakanywhere=true}
css: ../styles.css
---

```{r set-defaults, echo=FALSE, results=FALSE, message=FALSE}
knitr::opts_chunk$set(
  fig.dim = c(10, 8),
  out.width = "80%",
  eval = TRUE,
  echo = TRUE,
  results = "markup",
  message = FALSE
)
```

```{r load-r-helpers, echo=FALSE}
run_full <- TRUE
```

## The Analysis Goal

This vignette follows the procedure described in Zacharias et al. (2012) to
classify AKI vs non-AKI patients based on 24h urine NMR spectra. The goal is to
reproduce the reported performance of 76% accuracy and 83% AUC.

## The Original Analysis Procedure

Relevant parts from the original paper are:

*The region from 9.5 to 0.5 ppm of the 1D spectra was evenly split into bins of
0.01. After exclusion of the solvent area and the area containing the broad
urea signal from 6.5 to 4.5 ppm, 701 bins remained for each 1D spectrum.*

*All spectra were normalized prior to analysis by Quantile Normalization
(Bolstad et al. 2003; Kohl et al. 2012) to achieve equal distributions of
feature intensities across spectra.*

*Classification of spectra was performed using a SVM provided in the R-library
e1071 (http://cran.r-project.org/ web/packages/e1071) within a nested
cross-validation approach.*

*For this contribution a SVM with a radial basis function kernel was used
within a leave-five-out cross-validation.*

*The internal parameters of the SVM were calibrated within an inner
cross-validation loop employing a grid search procedure.*

## Helper Functions

Based on these instructions we're gonna implement the following helper functions:

`bin_spectra(spectra, regions, binwidth)`\
For creating a matrix of binned features from a list of spectra. Binning will be
done by splitting the given `regions` into bins of width `binwidth` and summing
the intensities of all data points that fall into each bin. The resulting binned
features will be stored in a matrix, where rows correspond to spectra and
columns correspond to bins.

`get_quantile_reference(X)`\
For computing the reference distribution for quantile normalization from a
feature matrix `X`.

`quantile_normalize(X, ref)`\
For applying quantile normalization to a feature matrix `X` using a reference
distribution `ref`.

`train_model(spectra, cls, cost, gamma, nfeat, normalize)`\
For training a radial SVM classifier with margin regularization strength
`cost`, RBF kernel width `gamma` and `nfeat` number of features. Features are
obtained by binning the given input spectra using `bin_spectra()` and
selecting the top `nfeat` based on t-scores from a Welch t-test, comparing the
two classes in `cls`. If `normalize` is TRUE, quantile normalization is
applied to the binned features before feature selection and SVM training. The
feature-selection based on t-scores is described in 'Gronwald et al., 2011'.
The returned object will contain the trained SVM model, the
quantile-normalisation reference distribution and the indices of the selected
features.

`find_best_params(X, y, costs, gammas, nfeats, normalize, k)`\
For training multiple SVMs across a grid using k-fold cross-validation to
select the best parameters. In this vignette we use `costs = (1:10) / 10`,
`gammas = 2^(-12:-5)` and `nfeats = 2:3`. We limit the number of features to
such a narrow range, because the original paper states, that the average
optimum is 2.4 across validation folds with a standard deviation of 0.5. This
means, that only values of 2 and 3 can have been observed (e.g. 2,2,2,3,3) or
the SD would have been higher. Since the goal of this article is not to rerun
the whole analysis workflow, but to reproduce the reported results as closely
as possible, there is no need to search a wider range of feature counts.

## The Final Classification Pipeline

The final classification pipeline will then be implemented as follows:

```{r pipeline-cv-cleaned}
benchmark_aki_classification_cv_cleaned <- function() {
  # Read spectra and metadata
  path <- metabodecon::download_example_datasets()
  aki_path <- file.path(path, "bruker", "aki")
  spectra <- metabodecon::read_spectra(aki_path)
  meta <- read.csv(file.path(aki_path, "s_MTBLS24.txt"), sep = "\t")
  y_bin <- as.integer(
    meta$Factor.Value.Acute.Kidney.Injury. == "Acute Kidney Injury"
  )
  y <- factor(y_bin, levels = c(0, 1), labels = c("Control", "AKI"))

  # Bin spectra and prepare y vector for classification
  X <- metabodecon:::bin_spectra(spectra)

  # Run classification across 5 folds and collect results
  test_ids <- metabodecon:::get_test_ids(
    nfolds = 5,
    nsamples = nrow(meta),
    y = y_bin
  )
  res <- lapply(test_ids, benchmark_aki_classification_holdout, X, y_bin)

  # Format results as 5 x 5 table
  results <- do.call(rbind, res)
  colnames(results) <- c("cost", "gamma", "nfeat", "auc", "acc")
  rownames(results) <- paste0("Fold", 1:5)

  # Add means and standard deviations across folds as additional rows
  means <- colMeans(results)
  sds <- apply(results, 2, sd)
  rbind(results, Mean = means, SD = sds)
  results
}
```

where `benchmark_aki_classification_holdout()` is defined as follows:

```{r pipeline-holdout}
benchmark_aki_classification_holdout <- function(te, X, y_bin) {
  tr <- setdiff(seq_len(nrow(X)), te)
  costs <- (1:10) / 10
  gammas <- 2^(-12:-5)
  nfeats <- 2:3
  p <- metabodecon:::find_best_params_cached(
    X[tr, ], y_bin[tr], costs, gammas, nfeats, TRUE, 5
  )
  svm <- metabodecon:::train_model(
    X[tr, ], y_bin[tr], p$cost, p$gamma, p$nfeat
  )
  yhat <- metabodecon:::predict_svm(svm, X[te, , drop = FALSE])
  auc <- metabodecon:::auc(y_bin[te], yhat$prob)
  acc <- mean(y_bin[te] == yhat$cls)
  c(p$cost, p$gamma, p$nfeat, auc, acc)
}
```

## Step by Step Execution

To make the analysis as transparent as possible, we will now execute the above
pipeline step by step, visualizing intermediate results and outputs along the
way, before showing the final results at the end of the vignette.

### Downloading the Data

The first thing we need to do is to download the data, if not already available
locally. This is done via `download_example_datasets()`. The AKI dataset is then
available in the `bruker/aki` subdirectory of the downloaded data. After that we
will use metabodecon's `tree_preview()` function to visualize the directory
structure of the downloaded data.

```{r download, eval=run_full}
xds_dir <- metabodecon::download_example_datasets()
aki_dir <- file.path(xds_dir, "bruker", "aki")
metabodecon::tree_preview(aki_dir)
```

### Reading Spectra and Metadata

Now that we have the data, we can read the spectra and metadata using
`read_spectra()` and `read_csv()`, respectively. The spectra will be stored as
object of class `spectra`, while the metadata will be stored in a data frame.
For more details about the spectra class, see article
[Metabdecon-Classes](https://spang-lab.github.io/metabodecon/articles/Classes.html).
To better understand the dataset, we will first use a combination of `print()`
and `summary()` to get an overview of the spectra and metadata objects. Then we
will use `plot_spectrum()` on a small, randomly selected subset of spectra to
visualize the raw data.

```{r read, eval=run_full, fig.width=10, fig.height=8}
#| fig.cap: |
#|   <strong>Figure 1.</strong> Four representative AKI spectra
#|   (indices 25, 50, 75, 100) in the region from 4.5 to 1.5 ppm.
#|   The plot title of each panel shows the sample ID and class label.
spectra <- metabodecon::read_spectra(aki_dir)
print(object.size(spectra), units = "Mb")
summary(head(spectra, 4))

meta <- read.csv(file.path(aki_dir, "s_MTBLS24.txt"), sep = "\t")
meta <- meta[, c("Sample.Name", "Factor.Value.Acute.Kidney.Injury.")]
meta <- setNames(meta, c("SampleID", "Type"))
N <- nrow(meta)
meta[c(1:4, (N - 3):N), ]

sel <- c(25, 50, 75, 100) # two normal and two AKI spectra
plots <- withr::with_par(list(mfrow = c(2, 2)), lapply(sel, function(i) {
  main <- paste0(meta$SampleID[i], " (", meta$Type[i], ")")
  metabodecon::plot_spectrum(
    spectra[[i]],
    foc_rgn = c(4.5, 1.5),
    sub1 = list(main = main)
  )
}))
```

### Binning the data

We now create the feature matrix by binning. Each row corresponds to one
spectrum and each column corresponds to one bin. Following the paper, we use the
range from 9.5 to 0.5 ppm and exclude the broad urea/water region from 6.5 to
4.5 ppm.

```{r binning, eval=run_full}
regs <- rbind(c(9.5, 6.5), c(4.5, 0.5))
X <- metabodecon:::bin_spectra(spectra, regions = regs, binwidth = 0.01)
y_bin <- as.integer(meta$Type == "Acute Kidney Injury")
y <- factor(y_bin, levels = c(0, 1), labels = c("Control", "AKI"))

dim(X)
table(y)
```

### Quantile Normalization

Quantile normalization forces each sample to have the same empirical
distribution. We compute a reference distribution from all samples and map each
sample to it.

```{r qn, eval=run_full}
#| fig.cap: |
#|   <strong>Figure 2.</strong> Sorted intensity profiles before
#|   (left) and after (right) quantile normalization for five
#|   selected spectra. After normalization, the empirical
#|   distributions are aligned across samples.
ref_qn <- metabodecon:::get_quantile_reference(X)
X_qn <- metabodecon:::quantile_normalize(X, ref_qn)

idx <- c(1, 25, 50, 75, 100)
withr::with_par(list(mfrow = c(1, 2)), {
  matplot(
    t(apply(X[idx, ], 1, sort)),
    type = "l",
    lty = 1,
    xlab = "rank",
    ylab = "intensity",
    main = "before quantile norm"
  )
  matplot(
    t(apply(X_qn[idx, ], 1, sort)),
    type = "l",
    lty = 1,
    xlab = "rank",
    ylab = "intensity",
    main = "after quantile norm"
  )
})
```

### PQN Normalization

Another common normalization method is Probabilistic Quotient Normalization
(PQN), which is based on the idea of normalizing each spectrum by a reference
spectrum that represents the typical sample in the dataset. The reference
spectrum is usually computed as the median spectrum across all samples. PQN is
particularly useful when there are large variations in the overall intensity of
the spectra, which is often the case in metabolomics data. At the end of this
section you can see a comparison of the results obtained with quantile
normalization and PQN.

```{r pqn, eval=run_full}
#| fig.cap: |
#|   <strong>Figure 3.</strong> Comparison of sorted intensity
#|   profiles after quantile normalization (left) and after PQN
#|   normalization (right) for the same five spectra.
ref_pqn <- metabodecon:::get_pqn_reference(X)
X_pqn <- metabodecon:::pqn_normalize(X, ref_pqn)

withr::with_par(list(mfrow = c(1, 2)), {
  matplot(
    t(apply(X_qn[idx, ], 1, sort)),
    type = "l",
    lty = 1,
    xlab = "rank",
    ylab = "intensity",
    main = "quantile norm"
  )
  matplot(
    t(apply(X_pqn[idx, ], 1, sort)),
    type = "l",
    lty = 1,
    xlab = "rank",
    ylab = "intensity",
    main = "PQN"
  )
})
```

### Making Cross Validation Folds

We use nested 5x5 CV. Outer folds are used for unbiased performance estimation.
Inside each outer training split, we build inner folds for parameter tuning.

```{r folds, eval=run_full}
#| fig.cap: |
#|   <strong>Figure 4.</strong> Nested cross-validation split
#|   structure. The first panel shows the outer 5-fold split.
#|   The second panel shows inner validation folds for outer fold 1.
n <- nrow(X)
outer_te <- metabodecon:::get_test_ids(
  nfolds = 5,
  nsamples = n,
  seed = 1,
  y = y_bin
)
metabodecon:::plot_folds(outer_te, nsamples = n, main = "Outer 5-fold split")

te1 <- outer_te[[1]]
tr1 <- setdiff(seq_len(n), te1)
inner_rel <- metabodecon:::get_test_ids(
  nfolds = 5,
  nsamples = length(tr1),
  seed = 1,
  y = y_bin[tr1]
)
inner_te <- lapply(inner_rel, function(i) tr1[i])
metabodecon:::plot_folds(
  test_ids = list(te1),
  nsamples = n,
  val_ids = inner_te,
  main = "Inner folds inside outer fold 1"
)
```

### Hyperparameter Tuning

First, we train one model for fold 1.1 and inspect its predictions.

```{r tune-one, eval=run_full}
#| fig.cap: |
#|   <strong>Figure 5.</strong> Predicted AKI probabilities on one
#|   inner validation fold (fold 1.1). Points are colored by true
#|   class and the dashed line marks the 0.5 decision threshold.
ok <- sapply(inner_te, function(v) {
  tr <- setdiff(tr1, v)
  length(unique(y_bin[tr])) == 2 && length(unique(y_bin[v])) == 2
})
if (!any(ok)) stop("No inner split with both classes found.")
val11 <- inner_te[[which(ok)[1]]]
tr11 <- setdiff(tr1, val11)

fit11 <- metabodecon:::train_model(X[tr11, , drop = FALSE], y_bin[tr11],
  cost = 0.5, gamma = 2^-10, nfeat = 3,
  normalize = TRUE
)
pred11 <- metabodecon:::predict_svm(fit11, X[val11, , drop = FALSE])
auc11 <- metabodecon:::auc(y_bin[val11], pred11$prob)
acc11 <- mean(y_bin[val11] == pred11$cls)

c(auc = auc11, acc = acc11)
metabodecon:::plot_prob_scatter(pred11$prob, y_bin[val11], "Fold 1.1 predictions")
```

Now we evaluate the full grid for fold 1.1 and then average AUCs across all
inner folds of outer fold 1.

```{r tune-grid, eval=run_full}
#| fig.cap: |
#|   <strong>Figure 6.</strong> Mean inner-fold AUC surfaces for
#|   hyperparameter tuning at nfeat = 2 (left) and nfeat = 3
#|   (right), across cost and gamma.
costs <- (1:10) / 10
gammas <- 2^(-12:-5)
nfeats <- 2:3

G11 <- metabodecon:::score_grid_cached(
  X,
  y_bin,
  tr11,
  val11,
  costs,
  gammas,
  nfeats,
  TRUE
)

valid_inner <- inner_te[ok]
G1 <- metabodecon:::score_grid_cached(
  X, y_bin, setdiff(tr1, valid_inner[[1]]), valid_inner[[1]],
  costs, gammas, nfeats, TRUE
)
for (i in 2:length(valid_inner)) {
  vai <- valid_inner[[i]]
  tri <- setdiff(tr1, vai)
  Gi <- metabodecon:::score_grid_cached(
    X,
    y_bin,
    tri,
    vai,
    costs,
    gammas,
    nfeats,
    TRUE
  )
  G1$auc <- G1$auc + Gi$auc
}
G1$auc <- G1$auc / length(valid_inner)

best1 <- G1[which.max(G1$auc), ]
best1

zlim <- range(G1$auc, na.rm = TRUE)
withr::with_par(list(mfrow = c(1, 2)), {
  metabodecon:::plot_auc_grid(G1, nfeat = 2, zlim = zlim, main = "mean inner AUC (nfeat=2)")
  metabodecon:::plot_auc_grid(G1, nfeat = 3, zlim = zlim, main = "mean inner AUC (nfeat=3)")
})
```

In this vignette we follow the standard strategy: compute one AUC per inner
validation fold and average these AUCs across folds to select parameters.

### Performance Evaluation

Finally, we repeat the procedure for all outer folds, collect fold-wise metrics,
and pool all outer-fold predictions for one global AUC/accuracy estimate.

```{r outer-cv, eval=run_full}
#| fig.cap: |
#|   <strong>Figure 7.</strong> Pooled outer-fold AKI probabilities
#|   from nested CV. Each point is one sample; color encodes class.
#|   The dashed line indicates the 0.5 classification threshold.
eval_outer_fold <- function(te, Xm, y_bin) {
  tr <- setdiff(seq_len(nrow(Xm)), te)
  p <- metabodecon:::find_best_params_cached(
    Xm[tr, , drop = FALSE], y_bin[tr], costs, gammas, nfeats, TRUE, 5
  )
  fit <- metabodecon:::train_model(
    Xm[tr, , drop = FALSE], y_bin[tr],
    p$cost, p$gamma, p$nfeat, TRUE
  )
  pred <- metabodecon:::predict_svm(fit, Xm[te, , drop = FALSE])
  list(
    ids = te,
    prob = pred$prob,
    cls = pred$cls,
    perf = c(
      cost = p$cost, gamma = p$gamma, nfeat = p$nfeat,
      auc = metabodecon:::auc(y_bin[te], pred$prob),
      acc = mean(y_bin[te] == pred$cls)
    )
  )
}

outer_res <- lapply(outer_te, eval_outer_fold, Xm = X, y_bin = y_bin)

perf <- do.call(rbind, lapply(outer_res, function(x) x$perf))
rownames(perf) <- paste0("Fold", seq_len(nrow(perf)))
perf <- rbind(perf, Mean = colMeans(perf), SD = apply(perf, 2, sd))
print(round(perf, 4))

prob_all <- rep(NA_real_, n)
for (i in seq_along(outer_res)) {
  prob_all[outer_res[[i]]$ids] <- outer_res[[i]]$prob
}

auc_all <- metabodecon:::auc(y_bin, prob_all)
acc_all <- mean((prob_all >= 0.5) == y_bin)
c(global_auc = auc_all, global_acc = acc_all)

metabodecon:::plot_prob_scatter(prob_all, y_bin, "Outer-fold pooled predictions")
```

The resulting values can now be compared to the reported 83% AUC and 76%
accuracy from the original paper.

## Extending the original results

We now rerun the full nested CV benchmark for two model types (`svm`,
`lasso`) and six normalization choices (`quantile`, `pqn`, `creatinine`,
`sum`, `median`, `none`).

The benchmarking helper supports the normalization timepoint through
`norm_time`:

- `norm_time = "cv"`: normalize correctly inside each training split.
- `norm_time = "global"`: normalize once before CV, then use `none` inside CV.

The table reports mean outer-fold AUC ± SD for both settings.

```{r extending-results, eval=run_full}
models <- c("svm", "lasso")
norms <- c("quantile", "pqn", "creatinine", "sum", "median", "none")
costs <- (1:10) / 10
gammas <- 2^(-12:-5)
nfeats <- 2:3

ext <- metabodecon:::benchmark_extension_table(
  X = X,
  y = y_bin,
  models = models,
  norms = norms,
  k = 5,
  costs = costs,
  gammas = gammas,
  nfeats = nfeats
)

tab <- data.frame(
  Method = paste(ext$model, ext$normalisation, sep = " + "),
  `AUC mean ± SD (CV-wise norm)` = ext$auc_cv_norm,
  `AUC mean ± SD (global pre-norm)` = ext$auc_global_norm,
  check.names = FALSE
)

print(tab, row.names = FALSE)
```