---
title: "AKI Original"
output:
  html_document:
    fig_caption: true
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{fontspec}
  - \usepackage{titling}
  - \pretitle{\begin{center}\LARGE\normalfont}
  - \posttitle{\par\end{center}\vskip 0.5em}
  - \usepackage{etoolbox}
  - \usepackage{fvextra}
  - \usepackage{needspace}
  # - \setmonofont{Menlo}
  - \definecolor{shadecolor}{RGB}{232,232,232}
  - \fvset{breaklines=true,breakanywhere=true}
  - \BeforeBeginEnvironment{Shaded}{\Needspace{10\baselineskip}}
  - \BeforeBeginEnvironment{Shaded}{\vspace{0.5em}}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines=true,breakanywhere=true,commandchars=\\\{\}}
  - \RecustomVerbatimEnvironment{verbatim}{Verbatim}{breaklines=true,breakanywhere=true}
css: ../styles.css
---

<style>
h1.title {
  font-weight: 400;
}
</style>

```{r set-defaults, echo=FALSE, results=FALSE, message=FALSE}
knitr::opts_chunk$set(
    out.width = "100%",
    fig.width = 10,
    fig.height = 4,
    cache = FALSE,
    cache.lazy = FALSE,
    eval = TRUE,
    echo = TRUE,
    results = "markup",
    message = FALSE
)
```

```{r load-r-helpers, echo=FALSE}
devtools::load_all(quiet = TRUE)
```

## The Analysis Goal

This vignette follows the procedure described in Zacharias et al. (2012) to
classify AKI vs non-AKI patients based on 24h urine NMR spectra. The goal is to
reproduce the reported performance of 76% accuracy and 83% AUC.

## The Original Analysis Procedure

Relevant parts from the original paper are:

*The region from 9.5 to 0.5 ppm of the 1D spectra was evenly split into bins of
0.01. After exclusion of the solvent area and the area containing the broad
urea signal from 6.5 to 4.5 ppm, 701 bins remained for each 1D spectrum.*

*All spectra were normalized prior to analysis by Quantile Normalization
(Bolstad et al. 2003; Kohl et al. 2012) to achieve equal distributions of
feature intensities across spectra.*

*Classification of spectra was performed using a SVM provided in the R-library
e1071 (http://cran.r-project.org/ web/packages/e1071) within a nested
cross-validation approach.*

*For this contribution a SVM with a radial basis function kernel was used
within a leave-five-out cross-validation.*

*The internal parameters of the SVM were calibrated within an inner
cross-validation loop employing a grid search procedure.*

## Step by Step Execution

To make the analysis as transparent as possible, we will execute the above
pipeline step by step over the course of this vignette, visualizing intermediate
results and outputs along the way, before showing the final results at the end
of the vignette.

## Downloading the Data

The first thing we need to do is to download the data, if not already available
locally. This is done via `download_example_datasets()`. The AKI dataset is then
available in the `bruker/aki` subdirectory of the downloaded data. After that we
will use metabodecon's `tree_preview()` function to visualize the directory
structure of the downloaded data.

```{r download}
path <- metabodecon::download_example_datasets()
aki_path <- file.path(path, "bruker", "aki")
metabodecon::tree_preview(aki_path)
```

## Reading Spectra and Metadata

Now that we have the data, we can read the spectra and metadata using
`read_spectra()` and `read_csv()`, respectively. The spectra will be stored as
object of class `spectra`, while the metadata will be stored in a data frame.
For more details about the spectra class, see article
[Metabdecon-Classes](https://spang-lab.github.io/metabodecon/articles/Classes.html).
To better understand the dataset, we will first use a combination of `print()`
and `summary()` to get an overview of the spectra and metadata objects. Then we
will use `plot_spectrum()` on a small, randomly selected subset of spectra to
visualize the raw data.

```{r read-meta}
meta <- read.csv(file.path(aki_path, "s_MTBLS24.txt"), sep = "\t")
meta <- meta[, c("Sample.Name", "Factor.Value.Acute.Kidney.Injury.")]
meta <- setNames(meta, c("SampleID", "Type"))
meta$Type <- ifelse(meta$Type == "Acute Kidney Injury", "AKI", "Control")
headtail(meta, 3)
```

```{r read-spectra}
spectra <- metabodecon::read_spectra(aki_path)
print(object.size(spectra), units = "Mb")
summary(head(spectra, 4))
```

```{r visualize-spectra}
#| fig.cap: |
#|   Two representative spectra in the region from 4.5 to 1.5 ppm.
plot_spectra_panel(spectra, meta = meta)
```

## Ordering Spectra and Metadata

Before we can bring spectra and metadata into the same order, we need to check
whether the sample IDs are the same in spectra and metadata.

```{r check-sample-ids}
s <- names(spectra) # spectra names
m <- meta$SampleID # metadata names
diffdf <- data.frame(sort(s), sort(m), equ = (sort(s) == sort(m)))
print(diffdf[!diffdf$equ, ])
```

<!--
               sort.s.             sort.m.   equ
17 AKI_8_24_105_110816 AKI_8_24_105_110812 FALSE
18 AKI_8_24_106_110816 AKI_8_24_106_110812 FALSE
19 AKI_8_24_107_110816 AKI_8_24_107_110812 FALSE
20 AKI_8_24_108_110816 AKI_8_24_108_110812 FALSE
21 AKI_8_24_109_110816 AKI_8_24_109_110812 FALSE
23 AKI_8_24_110_110816 AKI_8_24_110_110812 FALSE
-->

It seems as if the spectra names have a different date (110816) than the
metadata names (110812), so let's compare the distribution of dates in both
datasets:

```{r compare-sample-dates}
sd <- sapply(strsplit(s, "_"), `[`, 5)
md <- sapply(strsplit(m, "_"), `[`, 5)
table(sd)
table(md)
```

The 110816 appears only in the spectra names, while the other dates (110722,
110729 and 110812) appears in both, spectra data and metadata. Therefore, I'm
guessing that the the 110816 in the spectra names is wrong. However, because
renaming and re-uploading the spectra files is much more work than just patching
the metadata names, we will for now just patch the metadata names.

```{r patch-names}
meta$SampleID[meta$SampleID == "AKI_8_24_105_110812"] <- "AKI_8_24_105_110816"
meta$SampleID[meta$SampleID == "AKI_8_24_106_110812"] <- "AKI_8_24_106_110816"
meta$SampleID[meta$SampleID == "AKI_8_24_107_110812"] <- "AKI_8_24_107_110816"
meta$SampleID[meta$SampleID == "AKI_8_24_108_110812"] <- "AKI_8_24_108_110816"
meta$SampleID[meta$SampleID == "AKI_8_24_109_110812"] <- "AKI_8_24_109_110816"
meta$SampleID[meta$SampleID == "AKI_8_24_110_110812"] <- "AKI_8_24_110_110816"
```

Now let's see if we can finally map them:

```{r map-spectra-meta}
s <- names(spectra) # spectra names
m <- meta$SampleID # metadata names
all.equal(sort(s), sort(m))
```

Ok, we can map every meta row to a spectrum, so let's do that. There are many ways to do that, e.g.:

- Sort metadata in the same order as spectra
- Sort spectra in the same order as metadata
- Sort both of the by sample ID
- Sort both of them by sample number

For now, because it's easiest, I will sort metadata in the same way as spectra.

```{r read-meta-mbl}
rownames(meta) <- meta$SampleID
meta <- meta[names(spectra), ]
```

## Creatinine Normalization

According to the Masterthesis, accompanying the original paper, all spectra were
normalized by dividing through the creatinine signal at 3.05 ppm. The exact
quote from the Masterthesis is: 'The spectra were scaled to the reference region
of creatinine, 3.053 ppm - 3.011 ppm'. We can do this, by iterating over each spectrum object,
finding the datapoints in that region, summing up the intensities in that region to get the creatinine peak intensity,
and then dividing all intensities in the spectrum by that value.

```{r creatinine-normalization}
cr <- c(3.011, 3.053) # creatinine region
spectra_normed <- spectra # copy spectra to new object for normalization
for (i in seq_along(spectra)) {
    s <- spectra[[i]]
    idx <- which(s$cs >= cr[1] & s$cs <= cr[2])
    ci <- sum(s$si[idx]) # creatinine intensity
    spectra_normed[[i]]$si <- s$si / ci
}
```

## Binning the data

We now create the feature matrix by binning. Each row corresponds to one
spectrum and each column corresponds to one bin. Following the paper, we use the
range from 9.5 to 0.5 ppm and exclude the urea/water region from 6.5 to 4.5 ppm.
To counter the summing up of many huge intensity values, we divide the binned
values by 1e6 (one million) to get them into a nice, readable range.

```{r binning}
regions <- rbind(c(9.5, 6.5), c(4.5, 0.5))
X <- bin_spectra(spectra_normed, regions = regions, binwidth = 0.01)
which(!rownames(X) %in% meta$SampleID)

rownames(X) <- meta$SampleID
y <- factor(meta$Type, levels = c("Control", "AKI"))
fmt <- "Features: %d\nSamples: %d (%d AKI, %d Control)\n"
msg <- sprintf(fmt, ncol(X), nrow(X), sum(y == "AKI"), sum(y == "Control"))
cat(msg)
```

## Quantile Normalization

Quantile normalization forces each sample to have the same empirical
distribution. We compute a reference distribution from all samples and map each
sample to it.

```{r qn-heatmap-complex, out.width = "50%"}
#| fig.cap: |
#|   Quantile-normalized data: left sample summaries on a symlog2 x-scale,
#|   right heatmap over ppm bins.
ref_qn <- get_quantile_reference(X)
X_qn <- quantile_normalize(X, ref_qn)
rownames(X_qn) <- rownames(X)
draw_complex_heatmap(X, title = "Creatine Normed", x_axis = "ppm", mar = c(5, 1, 2, 1))
draw_complex_heatmap(X_qn, title = "Creatine+Quantile Normed", x_axis = "ppm", mar = c(5, 1, 2, 1))
```

## Making Cross Validation Folds

We use nested 21x20 CV. Outer folds are used for unbiased performance
estimation. Inside each outer training split, we build inner folds for
parameter tuning.

```{r folds, fig.height = 4, fig.width = 12}
#| fig.cap: |
#|   Nested cross-validation split
#|   structure. The first panel shows the outer 21-fold split.
#|   The second panel shows inner validation folds for outer fold 1.
n <- nrow(X)
ids <- seq_len(n)
te <- get_test_ids(ids, nfolds = 21, seed = 1, y = y)
tr <- lapply(te, function(x) setdiff(seq_len(n), x))
```

## Lasso Performance Estimation

```{r performance-estimation-glmnet, fig.height = 3}
# Train models with hypterparameters selected from inner CV
set.seed(1)
cvobjs <- lapply(seq_along(te), function(i) {
    glmnet::cv.glmnet(
        x = X_qn[tr[[i]], ],
        y = y[tr[[i]]],
        alpha = 1,
        nfolds = 5,
        family = "binomial"
    )
})

# Collect predictions on test sets
Y <- lapply(seq_along(te), function(i) {
    sid <- rownames(X_qn)[te[[i]]]
    true <- y[te[[i]]]
    prob <- predict(cvobjs[[i]], X_qn[te[[i]], ], s = "lambda.min", type = "response")[, 1]
    pred <- predict(cvobjs[[i]], X_qn[te[[i]], ], s = "lambda.min", type = "class")[, 1]
    pred <- factor(pred, levels = levels(y))
    data.frame(sid, true, prob, pred)
})
Y <- rbindlist(Y)
Y <- Y[order(Y$true, Y$prob), ]
Y$SampleNr <- as.integer(sapply(strsplit(Y$sid, "_"), `[`, 4))

# Plot sid vs. Probabilities. Color by class. Annotate x-ticks with sid. Add horizontal line at 0.5.
plot_empty(
    xlim = c(1, nrow(Y)), ylim = c(0, 1),
    axes = TRUE, ylab = "Predicted probability",
    xlab = "Row Nr in Data.frame",
    main = "Lasso CV Predictions"
)
text(seq_along(Y$prob), Y$prob, labels = Y$SampleNr, col = as.integer(Y$true))

# Show accuracy and AUC
accuracy <- mean(Y$true == Y$pred)
auc <- pROC::auc(Y$true, Y$prob)
fmt <- "Accuracy: %.2f%%\nAUC: %.3f\n"
table(Y$true, Y$pred)
cat(sprintf(fmt, accuracy * 100, auc))
```

## Global T-tests

```{r global-t-tests}
tscores <- compute_t_scores(X_qn, y)
pvals <- 2 * pt(-abs(tscores), df = nrow(X_qn) - 2)
withr::with_par(
    new = list(mfrow = c(1, 2)),
    code = {
        hist(tscores, breaks = 100, main = "T-scores for AKI vs Control", xlab = "T-score")
        hist(pvals, breaks = 100, main = "P-values for AKI vs Control", xlab = "P-value")
    }
)
df <- data.frame(feature = colnames(X_qn), tscore = tscores, pval = pvals)
df <- df[order(df$pval), ]
head(df, 20)
top2_fixed <- df$feature[1:2]
```

### SVM Performance Evaluation

```{r svm-hyperparams}
set.seed(1)
te <- get_test_ids(seq_len(nrow(X_qn)), nfolds = 5, seed = 1, y = y)
Y <- estimate_svm_performance(
    X = X_qn,
    y = y,
    test_ids = te,
    costs = 2^(0:5),
    gammas = 2^(-8:-3),
    nfeats = 2:3,
    inner_nfolds = 5,
    ncores = 5,
    verbose = FALSE
)
Yord <- Y[order(Y$true, Y$prob), ]
plot_empty(
    xlim = c(1, nrow(Yord)), ylim = c(0, 1),
    axes = TRUE, ylab = "Predicted score",
    xlab = "Row Nr in Data.frame",
    main = "Lasso CV Predictions"
)
text(
    x = seq_along(Yord$prob),
    y = Yord$prob,
    labels = as.character(Yord$sample),
    col = as.integer(Yord$true == "AKI") + 1
)

# Show accuracy and AUC
accuracy <- mean(Y$true == Y$pred)
auc <- pROC::auc(Y$true, Y$prob)
fmt <- "Accuracy: %.2f%%\nAUC: %.3f\n"
table(Y$true, Y$pred)
cat(sprintf(fmt, accuracy * 100, auc))
```

The resulting values can now be compared to the reported 83% AUC and 76%
accuracy from the original paper.
